---
output:
  rmarkdown::github_document:
    html_preview: no
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.path = "man/figures/README-", fig.align = "center",
                      fig.height = 4.5, fig.width = 4.5)
```

<!--- README.md is generated from README.Rmd. Please edit that file -->

## restopr: Interface to the 'Restopt' Ecological Restoration Planning Software

[![lifecycle](https://img.shields.io/badge/Lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![R-CMD-check-Ubuntu](https://img.shields.io/github/workflow/status/dimitri-justeau/restoptr/Ubuntu/master.svg?label=Ubuntu)](https://github.com/dimitri-justeau/restoptr/actions)
[![R-CMD-check-Windows](https://img.shields.io/github/workflow/status/dimitri-justeau/restoptr/Windows/master.svg?label=Windows)](https://github.com/dimitri-justeau/restoptr/actions)
[![R-CMD-check-MacOS](https://img.shields.io/github/workflow/status/dimitri-justeau/restoptr/macOS/master.svg?label=macOS)](https://github.com/dimitri-justeau/restoptr/actions)
[![Coverage Status](https://codecov.io/github/dimitri-justeau/restoptr/coverage.svg?branch=master)](https://app.codecov.io/gh/dimitri-justeau/restoptr)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/restoptr)](https://github.com/dimitri-justeau/restoptr)

```{r, include = FALSE}
devtools::load_all()
```

## Overview

Interface to the `restopt` ecological restoration planning software, which is specifically designed to identify connected, accessible, and compact areas for ecological restoration, with an emphasis on reducing habitat fragmentation and increasing habitat connectivity. This software reproduces the methodology described in [this article](https://www.researchgate.net/publication/346597935_Constrained_optimization_of_landscape_indices_in_conservation_planning_to_support_ecological_restoration_in_New_Caledonia), and is based on Constraint Programming (CP), which is a constrained optimization solving technique based on automated reasoning. Specifically, our model was implemented using [Choco-solver](https://choco-solver.org/), an open-source and state-of-the-art Java CP solver.

## Installation

### Package installation

The latest developmental version of the _restoptr R_ package can be installed using the following _R_ code.

```{r, eval = FALSE}
if (!require(remotes)) install.packages("remotes")
remotes::install_github("dimitri-justeau/restoptr")
```

Or with _devtools_:

```{r, eval = FALSE}
if (!require(devtools)) install.packages("devtools")
devtools::install_github("dimitri-justeau/rflsgen")
```


### System dependencies

The packages requires a Java Runtime Environment (JRE), version 8 or higher. Below we provide platform-specific instructions to install it.

#### _Windows_

Please install the latest Java Development Kit from [Oracle](www.oracle.com) website. To achieve this, navigate to the [downloads section of the website](https://www.oracle.com/java/technologies/javase-downloads.html), select the tab for the Windows operating system, and then download the x64 Installer file. After downloading the file, please run installer to install Java on your system. You will also need to ensure that the `PATH` environmental variable if configured so that _R_ can access Java. _restoptr_ relies on _rJava_ for the communication between _R_ and _Java_. If you have any trouble during the installation of _restopt_ due to _rJava_, please refer to _rJava_'s documentation: https://rforge.net/rJava/index.html.

#### _Ubuntu_

For recent versions of Ubuntu (18.04 and later), the Java libraries are available through official repositories. They can be installed using the following system commands.

```{bash, eval = FALSE}
sudo apt-get install default-jdk
```

If you want to install a specific JRE version, please follow instructions from [Oracle](https://www.oracle.com/java/technologies/javase-downloads.html), or [OpenJDK](https://openjdk.java.net/install/).

#### _Linux_

Please follow instructions from [Oracle](https://www.oracle.com/java/technologies/javase-downloads.html), or [OpenJDK](https://openjdk.java.net/install/).

#### _MacOS_

The easiest way to install the Java libraries is using [HomeBrew](https://brew.sh/). After installing HomeBrew, the Java libraries can be installed using the following system commands.

```{bash, eval = FALSE}
brew install openjdk
```

Please note that you might also need to ensure that the `PATH` environmental variable if configured so that _R_ can access Java.

## Citation

Please cite _restoptr_ when using it in publications.

> Justeau-Allaire, D., Vieilledent, G., Rinck, N., Vismara, P., Lorca, X., & Birnbaum, P. (2021). Constrained optimization of landscape indices in conservation planning to support ecological restoration in New Caledonia. Journal of Applied Ecology, 58(4), 744‑754.

This article describes the methodology. We will provide a specific citation for ressource the R package soon.

## Usage

The first thing to do to use _restoptr_ is to load the package:

```{r}
library(restoptr)
```

We will now create and solve a restoration optimization problem. Two input
rasters are necessary, the *existing_habitat* raster and the *restorable_habitat*
raster. **Important**: Both raster mus have the same dimensions and the same
spatial extent.

- The *existing_habitat* raster indicates the habitat area (raster value 1),
  the non-habitat area (raster value 0), and area which is not part of the
  landscape (raster value NA, or NODATA, e.g. ocean if the landscape is
  terrestrial).
  
- The *restorable_habitat* raster indicates the amount of habitat that can be
  restored for each planning unit (i.e. raster cells that are available for 
  restoration, i.e. non-habitat).
  
Example rasters, from the use case presented in [this study](https://www.researchgate.net/publication/346597935_Constrained_optimization_of_landscape_indices_in_conservation_planning_to_support_ecological_restoration_in_New_Caledonia)
are included in the package:

```{r, out.width="100%", fig.width = 10}
habitat_data <- rast(
  system.file("extdata", "habitat.tif", package = "restoptr")
)
restorable_data <- rast(
  system.file("extdata", "restorable.tif", package = "restoptr")
)
plot(rast(list(habitat_data, restorable_data)), nc = 2)
```

To instantiate a base restoration optimization problem from two input rasters,
use the `restopt_problem()` function:

```{r}
p <- restopt_problem(existing_habitat = habitat_data, 
                     restorable_habitat = restorable_data) 
```

Then, we can add constraints to this base problem. For instance, lets add a
locked-out constraint, to restrict the number of planning units that can be
selected for restoration. Such a constraint can be used to account for existing
land-use practices, feasibility of restoration activities, and stakeholder 
preferences.

```{r}
locked_out_data <- rast(
 system.file("extdata", "locked-out.tif", package = "restoptr")
)
p <- p %>% add_locked_out_constraint(data = locked_out_data)
```

We can also add a constraint on the amount of restored area that is allowed by the
project:

```{r}
p <- p %>% add_restorable_constraint(90, 220, 23)
```

And finally a compactness constraint, which limits the spatial extent of the
selected restoration area:

```{r}
p <- p %>% add_compactness_constraint(5)
```

Once we have added constraints to the problem, we need to define an optimization
objective. For example, lets configure `restopr` to identify, under the previous
constraints, which restoration areas maximises the effective mesh size (MESH).

```{r}
p <- p %>% set_max_mesh_objective()
```

*Note* The effective mesh size is a measure of landscape fragmentation based on
the probability that two randomly chosen points are located in the same patch
[Jaeger, 2000](https://doi.org/10.1023/A:1008129329289). Maximizing it in the context of restoration favours fewer and 
larger patches.

Finally, we use the `solve()` function to identify the optimal restoration area,
according to the constraints and the optimization objective.

```{r, out.width="100%", fig.width = 10}
s <- solve(p)
plot(
  x = s, main = "solution",
  col = c("#E5E5E5", "#fff1d6", "#b2df8a", "#1f78b4")
)
```

You can retrieve the attributes of the solution using the `attributes()` function:

```{r}
attributes(s)
```

## Getting help

If you have any questions about _restoptr_, improvement suggestions, or if you detect a bug, please [open an issue](https://github.com/dimitri-justeau/restoptr/issues/new/choose) in this GitHub repository.
