---
output:
  rmarkdown::github_document:
    html_preview: no
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.path = "man/figures/README-", fig.align = "center",
                      fig.height = 4.5, fig.width = 4.5)
```

<!--- README.md is generated from README.Rmd. Please edit that file -->

## restopr: Interface to the 'Restopt' Ecological Restoration Planning Software

[![lifecycle](https://img.shields.io/badge/Lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![R-CMD-check-Ubuntu](https://img.shields.io/github/workflow/status/dimitri-justeau/restoptr/Ubuntu/master.svg?label=Ubuntu)](https://github.com/dimitri-justeau/restoptr/actions)
[![R-CMD-check-Windows](https://img.shields.io/github/workflow/status/dimitri-justeau/restoptr/Windows/master.svg?label=Windows)](https://github.com/dimitri-justeau/restoptr/actions)
[![R-CMD-check-MacOS](https://img.shields.io/github/workflow/status/dimitri-justeau/restoptr/Mac%20OSX/master.svg?label=MacOS)](https://github.com/dimitri-justeau/restoptr/actions)
[![Coverage Status](https://codecov.io/github/dimitri-justeau/restoptr/coverage.svg?branch=master)](https://app.codecov.io/gh/dimitri-justeau/restoptr)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/restoptr)](https://github.com/dimitri-justeau/restoptr)

```{r, include = FALSE}
devtools::load_all()
```

## Overview

Interface to the `restopt` ecological restoration planning software, which is specifically designed to identify connected, accessible, and compact areas for ecological restoration, with an emphasis on reducing habitat fragmentation and increasing habitat connectivity.

## Installation

### Package installation

The latest developmental version of the _restoptr R_ package can be installed using the following _R_ code.

```{r, eval = FALSE}
if (!require(remotes)) install.packages("remotes")
remotes::install_github("dimitri-justeau/restoptr")
```

### System dependencies

The packages requires [Java (version 8 or higher)](https://www.java.com/). Below we provide platform-specific instructions to install it.

#### _Windows_

Please install the latest Java Development Kit from [Oracle](www.oracle.com) website. To achieve this, navigate to the [downloads section of the website](https://www.oracle.com/java/technologies/javase-downloads.html), select the tab for the Windows operating system, and then download the x64 Installer file. After downloading the file, please run installer to install Java on your system. You will also need to ensure that the `PATH` environmental variable if configured so that _R_ can access Java.

#### _Ubuntu_

For recent versions of Ubuntu (18.04 and later), the Java libraries are available through official repositories. They can be installed using the following system commands.

```{bash, eval = FALSE}
sudo apt-get install default-jdk
```

#### _Linux_

For Unix-alikes, `javac` (>= 8) is required.

#### _MacOS_

The easiest way to install the Java libraries is using [HomeBrew](https://brew.sh/). After installing HomeBrew, the Java libraries can be installed using the following system commands.

```{bash, eval = FALSE}
brew install openjdk
```

Please note that you might also need to ensure that the `PATH` environmental variable if configured so that _R_ can access Java.

## Citation

TODO.

## Usage

To reproduce the example from https://github.com/dimitri-justeau/restopt

### Example : maximize MESH

```{r, eval = FALSE}
library(restoptr)
habitat <- terra::rast("./inst/extdata/habitat.tif")
restorable <- terra::rast("./inst/extdata/restorable.tif")
accessible <- terra::rast("./inst/extdata/accessible.tif")
problem <- RestoptProblem(habitat=habitat, restorable=restorable, accessible=accessible)
problem <- postNbComponentsConstraint(problem, 1, 1)
problem <- postRestorableConstraint(problem, 90, 110, 23, 0.7)
problem <- postCompactnessConstraint(problem, 6)
result <- maximizeMESH(problem, 3)
```

### Example : maximize IIC

```{r, eval = FALSE}
library(restoptr)
habitat <- terra::rast("./inst/extdata/habitat.tif")
restorable <- terra::rast("./inst/extdata/restorable.tif")
accessible <- terra::rast("./inst/extdata/accessible.tif")
problem <- RestoptProblem(habitat=habitat, restorable=restorable, accessible=accessible)
problem <- postNbComponentsConstraint(problem, 1, 1)
problem <- postRestorableConstraint(problem, 90, 110, 23, 0.7)
problem <- postCompactnessConstraint(problem, 6)
result <- maximizeIIC(problem, 3)
```

## Getting help

TODO.
